---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cgcRelativities

<!-- badges: start -->
![GitHub R package version](https://img.shields.io/github/r-package/v/zerogetsamgow/cgcRelativities) 
![GitHub last commit](https://img.shields.io/github/last-commit/zerogetsamgow/cgcRelativities) 
<!-- badges: end -->

cgcRelativities provides tidy data versions of certain data tables from the Commonwealth Grant's Commission's annual [Reports to Government](https://www.cgc.gov.au/reports-for-government).

At present data on relativities and own source revenue are included in the package.

## Installation

You can install the development version of cgcRelativities from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("zerogetsamgow/cgcRelativities")
```

## Relativities 
Relativities data is stored in three files 
Data is stored as 

* `relativies_recommended.rda`, the actual relativities recommended by the CGC. 

```{r recommended}
recommended = cgcRelativities::relativities_recommended

dplyr::glimpse(recommended)
```

* `relativies_annual.rda`, the annual relativities based on one year of assessment date used to calculate the recommendations 
```{r annual}
annual = cgcRelativities::relativities_annual

dplyr::glimpse(annual)
```

`relativies_floorless.rda` a set of imputed relativities assumming 2018 reforms had not been implemented.
```{r floorless}
floorless = cgcRelativities::relativities_floorless

dplyr::glimpse(floorless)

```
