---
title: "submission_example"
author: "Sam Gow"
output: rmarkdown::html_vignette
vignette: >
  "%\\VignetteIndexEntry{submission_example}
  %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}\n"
fontsize: 11pt
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 600
)
library(cgcRelativities)
library(dplyr)
```

## Introduction

This submission responds to two questions asked by the Productivity Commission in its Issues paper.

-   Do the 2018 GST distribution reforms strike the right balance between responding to changing circumstances and providing certainty around revenue?

-   Have the GST distribution arrangements since the 2018 legislative changes delivered a reasonable level of horizontal fiscal equalisation?

In summary, this submission contends that the answer to both questions is 'no, the methods used to achieve horizontal fiscal equalisation could be improved'.

All analysis in this document can be accessed at <https://github.com/zerogetsamgow/cgcRelativities>. The R package available at this link provides the CGC data used in my analysis in tidy formats (and example code to tidy additional data contained in the CGC's update releases). Commission staff may find the data and code useful for their own analysis. 

### Do the 2018 GST distribution reforms strike the right balance between responding to changing circumstances and providing certainty around revenue?

In its [Issues Paper](https://assets.pc.gov.au/2025-11/gst-reforms-issues.pdf?VersionId=YQP9pGfxnDPTIkkKhn48nGko5OncLRWw), the Productivity Commission asked 'whether the [current approach to horizontal fiscal equalisation] strike[s] the right balance between responding to changing circumstances and providing certainty around revenue'.

This submission contends the CGC's attempts to smooth variation in GST payments over time undermine the objective of horizontal fiscal equalisation and by doing so risk creating mismatches between state and territory revenue and expenditure requirements. Horizontal fiscal equalisation should aim to match revenue with expenditure needs in a given year, rather than revenue from year to year. By doing the latter, it increases certainty about each state and territory's total revenue, rather than a single source, and better matches total revenue to expenditure needs.

The CGC's current approach involves three features that aim to provide additional certainty about each state and territory's GST payment revenue:

-   recommended relativities are provided prior to the commencement of the year they apply - by providing its recommendations for 2025-26 before the start of that year, the CGC reduces uncertainty about how much each state will receive when they are preparing their 2025-26 budgets (some uncertainty remain as the total amount of GST and the population estimates used to calculate payments won't be determined until the second half of 2026). However, by doing so the CGC must base its assessments of fiscal capacity and need on data further removed from the year they will apply.
-   the recommended relativities (prior to any application of a floor) average the CGC's assessment of state fiscal capacity in three previous years known as 'annual relativities'. Averaging over three years smoothes variations in the series of recommended relativities. For 2025-26, the recommended relativies averaged the CGC's assessment of state fiscal capacities in 2023-24, 2022-23, and 2021-22.
-   since 2022-23, the CGC's recommendation have limited by a floor initially set at 0.70 in 2022-23 then increased to 0.75 from 2024-25.[^1]

[^1]: This submission does not deal with the latter feature which is clearly flawed and is likely to be the subject of multiple submissions that will make strong arguments for its removal.

By making the GST payments more predictable in this way, the CGC is making them less effective at equalising fiscal capacity. However, the case for reducing uncertainty is unclear. State budgets are also affected by many other sources of uncertainty. State treasurer's may be more certain about the amount of GST payments they will receive due to these features, but overall this approach results in the risk of them not having the fiscal capacity in a given year to provide the average level of services. It also risks amplifying fluctuations in own source revenue.

For example, if a state's revenue raising capacity drops between the assessment period and the application year, the GST payment calculated using the CGC's current methodology will result in them having insufficient fiscal capacity to fund services at the average level. As such, increasing certainty about GST payments comes at the expense of equalising fiscal capacity and is likely to make budgeting harder, not easier.

This submission argues that increasing the predictability of GST payments to the states should be treated as a secondary consideration to the CGC's principal objective - fiscal equalisation - if not abandoned altogether.

Combined the first two features listed above result in CGC recommendations that can poorly reflect the fiscal capacity of states in the year they apply. This can be seen in the chart below, Western Australia's strong fiscal capacity through 2020-21 and 2021-22 wasn't reflected in the relativities recommended (yellow dots) for those years, and would have taken effect in 2022-23 if a floor had not been introduced. As a result, Western Australia received much more revenue than was required to meet its fiscal needs in 2020-21 and 2021-22 and anticipated receiving insufficient revenue to meet those needs in 2022-23. The anticipation of this shortfall likely inspired that state's argument for a floor.

### Different approaches to calculating relativities

```{r poor_predicter, echo = FALSE}

# Get last estimate of relativity for each year
relativies_actual =
  relativities_annual |> 
  dplyr::group_by(financial_year , state_name) |>
  dplyr::arrange(state_name, desc(financial_year ), desc(update_year)) |>
  dplyr::filter(dplyr::row_number() == dplyr::n()) |> 
  dplyr::mutate(relativity_type = "Actual")

# Get annual relativity for Prior-year
relativities_prior =
  relativities_annual |>
  dplyr::group_by(financial_year , state_name) |>
  dplyr::arrange(state_name, desc(financial_year ), desc(update_year)) |>
  dplyr::filter(dplyr::row_number()==n()) |>
  dplyr::mutate(financial_year = financial_year + lubridate::years(1)) |>
  dplyr::mutate(relativity_type="Prior year") |>
  dplyr::ungroup()


# Combine all data
all.relativities =
  dplyr::bind_rows(
    relativies_actual,
    relativities_prior,
    relativities_floorless |>
      dplyr::mutate(relativity_type = "Recommended")
  ) 

# Create plot for headline chart

ggplot2::ggplot(
  all.relativities |> 
    dplyr::filter(!stringr::str_detect(state_name,"Northern|Tasmania")),
  ggplot2::aes(
    x = financial_year - months(6),
    y = relativity,
    colour = relativity_type))+
  ggplot2::geom_point(size = .6, alpha=.8)+
  ggplot2::facet_wrap(
    facets = vars(state_name), nrow=2,
    labeller =
      ggplot2::labeller(
        state_name =
          ggplot2::label_wrap_gen(width = 20)),
    axes = "all_y") +
  ggplot2::scale_x_date(
    name=NULL,
    date_labels = "%y",
    date_breaks="1 years")+
  ggplot2::coord_cartesian(
   xlim = c(lubridate::dmy("30-9-2018"),NA)) +
  ggplot2::scale_y_continuous(
    name="Relativity")+
  ggplot2::scale_colour_manual(
    name=NULL, 
    values = c(
      vpstheme::bv.mint,
      vpstheme::bv.pink,
      vpstheme::bv.cerulean)) +
  vpstheme::theme_vps_dh(
    base_size = 20)+
  ggplot2::labs(
    caption = 
      "Tasmania and Northern Territory omitted for legibility.") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(hjust=0.5))
```

Fiscal equalisation could be improved if the CGC's recommendations:

-   applied in the year they are provided - that is, the recommendations made on 14 March 2025 would apply to 2024-25 payments.

-   were based only on an assessment of the most recent year data is available for, rather than an average of three years.

Effectively this would result in recommended relativities using only the most up to date actuals data available.

The chart above shows the relativities based on this approach as pink dots. In general, they are a better reflection of the fiscal capacity when they would apply then the current recommendations. This can be seen in the chart below, where the assessed fiscal capacity for the prior year has generally been closer to actual fiscal capacity in any given year than the CGC's recommendations.

### Fiscal capacity in the prior year is a better predictor than the CGC's recommendations

```{r difference_plot, echo = FALSE}

difference.relativities =
  all.relativities |>
  dplyr::filter(!financial_year>
                  lubridate::ymd("2024-6-30"),
                !financial_year<
                  lubridate::ymd("2019-6-30")
                ) |>
  dplyr::group_by(
    state_name, financial_year) |>
          dplyr::arrange(state_name,   financial_year,relativity_type) |>
  dplyr::mutate(relativity= abs(relativity-first(relativity))) |>
  dplyr::filter(!stringr::str_detect(relativity_type,"Actual"))


ggplot2::ggplot(
  difference.relativities |>
    dplyr::filter(
      !stringr::str_detect(state_name,"Northern|Tasmania")),
  ggplot2::aes(
    x = financial_year - months(6),
    y = relativity,
    fill = relativity_type))+
  ggplot2::geom_col(
  #  width = 150,
    position = ggplot2::position_dodge())+
  ggplot2::scale_x_date(
    name=NULL,
    date_labels = "%y", 
    date_breaks = "1 years") +
  ggplot2::coord_cartesian(
    xlim = c(lubridate::dmy("30-6-2018"),NA)) + 
  vpstheme::scale_y_continuvps(
    name="Absolute difference from actual capacity")+
  ggplot2::scale_fill_manual(
    name=NULL, 
    values = c(
      ggplot2::alpha(vpstheme::bv.pink,.7),
      ggplot2::alpha(vpstheme::bv.cerulean,.7))) +
  ggplot2::facet_wrap(facets=vars(state_name),nrow=2) +
  ggplot2::labs(
    caption = 
      "Tasmania and Northern Territory omitted for consistency.") +
  ggplot2::guides(fill = ggplot2::guide_legend(keywidth = .5)) +
  vpstheme::theme_vps_dh(
    base_size = 20)+
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(hjust=.5))
```

As shown in the table below, relativities calculated using this method would have been closer on average to actual fiscal requirements in the year of application than the current approach for most states and territories over the period 2019-20 to 2024-25.

### Mean difference between fiscal need and recommended relativity using new method and as recommended, 2019-20 to 2024-25

```{r difference_table, echo = FALSE}

difference_summary =
  difference.relativities |> 
  dplyr::filter(
    financial_year>lubridate::dmy("30-6-2018")) |> 
  dplyr::group_by(state_name, relativity_type) |> 
  dplyr::summarise(mean_value = round(mean(relativity),5),.groups = 'drop') |>
  tidyr::pivot_wider(
    names_from = relativity_type, values_from = mean_value)

knitr::kable(
  difference_summary |> 
    dplyr::rename("State or territory" = "state_name",
           "New method"= "Prior year")) |> 
  kableExtra::kable_styling() 

```

If I was a state treasurer, I would prefer this outcome to the certainty provided by the current approach as it would help to ensure the state has the fiscal capacity to provide services at the average level. There are other options for states to smooth revenue over time (borrowing and saving) or better plan for changes (improving forecasting methods) that don't detract from fiscal equalisation.

There is also no evidence that increasing certainty of GST payments leads to better fiscal decision making. The evidence available suggests states and territory may spend [windfall payments](https://www.theage.com.au/national/victoria-s-spending-decision-reignites-spat-between-states-over-gst-carve-up-20250522-p5m1fs.html) or cut spending in the face of [shortfalls](https://www.abc.net.au/news/2024-04-22/nsw-treasurer-daniel-mookhey-warns-budget-shock-11b-gst-hole/103751654).

## Have the GST distribution arrangements since the 2018 legislative changes delivered a reasonable level of horizontal fiscal equalisation?

The current approach to assessing state own-source revenue is not delivering a reasonable level of horizontal fiscal equalisation. Replacing the current approach (aggregating assessments of individual own-source revenue streams) with a simpler measure (like state final demand) would be more efficient both in terms of administrative effort required and the tax system that might arise from removing disincentives for tax reform.

The Commonwealth Grants Commission calculates revenue and expenditure ‘needs’ to determine the relativity that is applied to state populations to determine their share of GST.

These needs needs reflects each state’s capacity to raise revenue and the amount each would spend to deliver similar levels of service (as well as assessments of borrowing and investment costs and Commonwealth payments other than the GST). A summary of these calculations for 2023-24 is shown below.

### Summary of CGC assessment calculations for 2023-24

```{r assessment_summary, echo = FALSE}
knitr::kable(
  latest_summary |>
  tidyr::pivot_wider(names_from = state_name) |> 
    mutate_if(
      is.numeric, 
      format, 
      scientific = FALSE,
      digits = 2, nsmall = 2) |> 
    dplyr::filter(financial_year == max(financial_year)) |> 
    dplyr::select(-financial_year) |> 
    dplyr::rename_with(
      \(x) stringr::str_replace(x,"_", " "))) |>
  kableExtra::kable_styling()

```

The importance of revenue capacity can be seen in Western Australia’s assessment. They were assessed as being able to raise nearly \$3 000 per person more than any other state. Compare this to the amount of GST per person to be shared, and we see much of Western Australia’s relative fiscal strength came from revenue.

Western Australia’s fiscal strength resulted in a decision to place a floor under each states relativity. Instead of imposing a floor, another option to address Western Australia’s declining GST share could have been to make sure the CGC's assessments better reflected capacity. The ‘assessed’ that prefaces each category in the table above is an important qualifier. GST relativities are based on the CGC’s assessment of each state’s capacity to raise revenue, not the amounts they raise.

I, and others, consider the current assessments to be flawed. The reasons for this view can be seen by comparing the CGC’s assessment of payroll tax revenue with actual revenue as shown below.

### Payroll tax revenue per capita, actual versus assessed

```{r payroll, echo = FALSE}
payroll_per_capita =
  dplyr::bind_rows(
    taxation_gfs |> 
     dplyr::filter(stringr::str_detect(revenue_name,"Payroll")),
    revenue_assessed |> 
      dplyr::filter(stringr::str_detect(revenue_name,"Payroll")) |> 
      dplyr::group_by(
        financial_year, state_name, revenue_name, revenue_type) |> 
      dplyr::arrange(
        financial_year, state_name, revenue_name, revenue_type, desc(update_year)) |> 
      dplyr::slice_head(n=1)) |> 
  dplyr::left_join(population_erp, by = join_by("state_name","financial_year")) |>
  dplyr::mutate(
    per_capita = (1e6*revenue)/population
  )

#PLot
ggplot2::ggplot(
  data = 
    payroll_per_capita  |>
    dplyr::filter(!stringr::str_detect(state_name,"Northern|Tasmania")),
    ggplot2::aes(
      x = financial_year,
      y = per_capita,
      colour = state_name,
      linetype = revenue_type)
    ) + 
  ggplot2::geom_line(linewidth = .5) +
   ggplot2::scale_linetype_manual(
     name = NULL,
     values = c("Actual"=1,"Assessed"=2))+
   ggplot2::facet_wrap(
     facets = vars(state_name), nrow=2,
     labeller = 
       ggplot2::labeller(
         state_name = 
           ggplot2::label_wrap_gen(width = 20)),
     axes = "all_y") +
  ggplot2::scale_colour_manual(
    guide = 'none',
    values=strayr::palette_state_name_2016,name=NULL)+
  ggplot2::scale_x_date(
    name=NULL,
    date_labels = "%y",
    date_breaks="1 years",
    )+
  ggplot2::coord_cartesian(
    xlim = c(lubridate::dmy("30-9-2018"),NA),
    ylim = c(0,2000)) +
  vpstheme::scale_y_continuvps(
    name="Revenue per capita, $ 000s", 
    labels = scales::label_number(scale = 1/1e3))+
  ggplot2::guides(linetype = ggplot2::guide_legend(keywidth = 2)) +
  vpstheme::theme_vps_dh(base_size = 20)
```

Your gaze might be drawn to the ACT in this chart. Not surprisingly. The ACT raises 50 per cent more revenue per person from payroll tax than the CGC assessment1. The gap between the ACT’s actual and assessed revenue reflects the assessment assuming limits on the ACT imposing payroll tax on the largest employer in the capital, the Australian Government, reduce their capacity. As a result of these limits, the CGC considers the ACT to have a similar or smaller payroll tax base per person than most other states.

So, why is the ACT’s actual revenue so much higher? Defenders of the current approach would suggest it results from a policy choice to impose a higher tax rate, not capacity. The CGC’s methodology avoids penalising jurisdictions for such choices, but, while the ACT does impose the highest payroll tax rate, this policy choice likely reflects capacity not captured in the current methodology.

While payroll tax is legally imposed on employers, it is generally accepted that the economic incidence of the tax falls on workers (in the form of lower wages) and/or consumers (in the form of higher prices). ACT employers likely can’t pass the tax on to workers (they are competing with the Australian Government who don’t pay the tax after all), so it is more likely passed on to consumers in the form of higher prices.

Most payroll tax in the ACT is raised from large national businesses (like the major supermarkets, banks and the big four consulting firms) who pass the cost on to ACT consumers, both residents and other businesses (including the Australian Government). They can do this as their customers have the capacity to pay - ACT residents have the highest average incomes in the country - and likely limited choice but to do so. The current methodology by looking at the size of the legal tax base fails to reflect the capacity of those who bear the cost of the tax.

So, perhaps, rather than assessing capacity to raise revenue by looking at the size of existing tax bases, the CGC could be looking at other measures - like the earnings of residents. I am not the first to make this observation. This idea first came to my attention from Peter Abelson more than a decade ago. Peter later brought raised them in his [submission to the Productivity Commission’s inquiry into horizontal fiscal equalisation](https://assets.pc.gov.au/__data/assets/pdf_file/0005/218489/sub009-horizontal-fiscal-equalisation.pdf). Peter argued state ‘revenue raising capacity is primarily a function of the real household disposable income of residents after allowances for major cost of living differences’. At the time, assessments based on this measure would lead to much higher assessments for the ACT and Victoria and much lower estimates for Queensland and Western Australia.

Apart from its novelty, I suspect a reason Peter’s very good idea failed to capture the attention of the PC and the CGC is that it would have required replacing an accepted measure with one that would require new effort to measure (and even more effort to agree on how it should be measured).

As a practitioner, I have always looked for simple solutions to challenging data problems. For example, I was able to improve forecasts of GST relativities by eschewing the normal data-rich option (which requires data on every state’s revenue from each tax) and instead using the smallest possible set of data of past relativities and select forecasts. Similarly, I can see how Peter’s idea could be implemented using available data contained in the state accounts prepared by the Australian Bureau of Statistics.

```{r revenue_capacity, include = FALSE}
# Assessed revenue data and clean to get total per capita
assessed_revenue =
  cgcRelativities::revenue_assessed |>
  
  group_by(state_name, financial_year, revenue_type) |> 
  filter(stringr::str_detect(revenue_name,"Total")) |>
  arrange(state_name, financial_year, revenue_type, desc(update_year)) |> 
  slice_head(n = 1) |> 
  mutate(revenue = 1e6*revenue) |> 
  left_join(
    cgcRelativities::population_erp
  ) |> 
  ungroup() |> 
  mutate(capacity = revenue/population,
         need = sum(revenue)/sum(population) - capacity)

# Construct actual revenue per capita based on GFS
cgc_revenue =
  cgcRelativities::revenue_gfs |> 
  filter(stringr::str_detect(gfs_subcategory,"Interest|Dividend|Taxation|Other|Sales|Royalty")) |> 
  group_by(state_name, financial_year) |> 
  summarise(revenue = 1e6*sum(value),
            revenue_type = "Actual",
            .groups = "drop") 

# Calculate macro based capacity and needs
macro_needs =
  cgc_revenue |> 
  left_join(
    cgcRelativities::estimated_gsp,
    by = join_by("state_name", "financial_year")
  ) |> 
  left_join(
    cgcRelativities::population_erp,
    by = join_by("state_name", "financial_year")
  ) |> 
  mutate(
    gsp_per_capita = gsp/population,
    rev_per_capita = revenue/population
  ) |> 
  group_by(financial_year) |> 
  mutate(
    gsp_per_aus = sum(gsp)/sum(population),
    rev_per_aus = sum(revenue)/sum(population)
  ) |> 
  ungroup() |> 
  select(
   state_name, financial_year, population, contains("per")
  ) |> 
  mutate(
    capacity = rev_per_aus/gsp_per_aus*gsp_per_capita,
    need = rev_per_aus - capacity
  )

# Combine data for comparision plots
capacity_comparison =
  bind_rows(
     assessed_revenue |> 
     select(state_name, financial_year, capacity, need) |> 
     mutate(name = "Current"),
     macro_needs |> 
      select(state_name, financial_year, capacity, need) |> 
      mutate(name = "Proposed") 
     )
```

### State and territory capacity to raise revenue under different methods

```{r capacity_plot, echo = FALSE}

ggplot2::ggplot(
  data = capacity_comparison,
  ggplot2::aes(
    x = financial_year,
    y = capacity, 
    colour = state_name,
    linetype = name
  )) +
 ggplot2::geom_line(linewidth = .5) +
  ggplot2::scale_linetype_manual(
     name = NULL,
     values = c("Current"=1,"Proposed"=2))+
   ggplot2::facet_wrap(
     facets= vars(state_name),
     nrow=2,
     labeller = 
       ggplot2::labeller(
         state_name = 
                   ggplot2::label_wrap_gen(width = 20)),
             axes = "all_y") +
  ggplot2::scale_colour_manual(
    guide = 'none',
    values=strayr::palette_state_name_2016,name=NULL)+
  ggplot2::scale_x_date(
    name=NULL,
    date_labels = "%y",
    date_breaks="1 years")+
  vpstheme::scale_y_continuvps(
    name="Revenue per capita, $ 000s", 
    labels = scales::label_number(scale = 1/1e3))+
  ggplot2::coord_cartesian(
    ylim = c(0,15000), 
    xlim = c(lubridate::dmy("30-9-2018"),NA)) +
  ggplot2::guides(linetype = ggplot2::guide_legend(keywidth = 2)) +
  vpstheme::theme_vps_dh(base_size = 20) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(hjust=0.5))

```

The chart aboveshows the CGC’s assessment of revenue raising capacity compared to a proposal to assess revenue raising capacity using only state final demand. This model assumes each state’s capacity to raise revenue is based on the size of its economy. Their capacity is then the average state revenue per dollar of gross domestic product per person multiplied by state final demand. Consistent with the discussion above, the ACT’s assessed capacity is significantly increased, as is the capacity of the Northern Territory and to a lesser extent the capacities of Victoria, South Australia and Tasmania.

The revenue ‘needs’ that arise from these two options are shown below. The changes in relative capacity results in Western Australia’s share of GST being less affected by its revenue raising capacity, though it would still be a donor state. The amounts received by the recipient state’s would also be reduced, and in the case of the territories their revenue raising capacity would result in them going from requiring additional GST payments to make up for a lack of revenue raising capacity to being donors (though the Northern Territory’s expenditure requirements would still see it receive much more than an equal per capita share of GST).

### Comparison of GST revenue needs based on different revenue assessment methods

```{r revenue_need, echo = FALSE}
ggplot2::ggplot(
  data = capacity_comparison,
  ggplot2::aes(
    x = financial_year,
    y = need, 
    colour = state_name,
    linetype = name)) +
 ggplot2::geom_line(linewidth = .5) +
  ggplot2::geom_hline(
    yintercept = 0,
    colour = vpstheme::bv.charcoal, 
    linewidth = .4)+
  ggplot2::scale_linetype_manual(
     name = NULL,
     values = c("Current"=1,"Proposed"=2))+
   ggplot2::facet_wrap(
     facets = vars(state_name), nrow=2,
     labeller = 
       ggplot2::labeller(
         state_name = 
           ggplot2::label_wrap_gen(width = 20)),
     axes = "all_y") +
  ggplot2::scale_colour_manual(
    guide = 'none',
    values = strayr::palette_state_name_2016,name=NULL)+
  ggplot2::scale_x_date(
    name = NULL,
    date_labels = "%y",
    date_breaks = "1 years",
    )+
  vpstheme::scale_y_continuvps(
    name="Revenue per capita, $000s", 
    labels = scales::label_number(scale = 1/1e3))+
  ggplot2::coord_cartesian(
    xlim = c(lubridate::dmy("30-9-2018"),NA),
    ylim = c(-5.5e3,5e3) )+
  ggplot2::guides(linetype = ggplot2::guide_legend(keywidth = 2)) +
  vpstheme::theme_vps_dh(base_size = 20) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(hjust=0.5))
```

The results of adopting this proposed approach, or any equivalent improvement to assessing revenue raising capacity, would have reduced the need to impose a floor under the GST relativities. Assessed state fiscal capacities would have been much more consistent over time and, I would argue, much more reflective of actual capacity.

Finally, one other benefit of adopting an approach to assessing revenue raising capacity based on a macro measure like state final demand is that it would eliminate disincentives to tax reform that exist in the current approach. Abolition of stamp duty is seen as an important economic reform. However, the current assessment methodology discourages individual states from abolishing inefficient taxes as they would continue to be assessed as having capacity to raise revenue from the stamp duty base and risk their assessed capacity increasing as transactions of properties would increase in volume without the tax.

The final chart in this note shows assessed and actual stamp duty revenue, this includes stamp duty on conveyances as well as other stamp duties. It shows the effect of the current method in the assessment of the Australian Capital Territory's revenue exceeding actual revenue. The Australian Capital Territory is in the process of abolishing stamp duty, but this is ignored by the CGC. Assessments of revenue raising capacity based on a macro measures like that proposed above would largely overcome this issue (there would still be a smaller disincentive to the extent that state final demand per capita increased). To the extent that it does, then it may help to unlock some barriers to productivity enhancing tax reform.

### Stamp duty revenue per capita, actual versus assessed

```{r stamp_duty, echo = FALSE}
# Get data
stamp_per_capita =
  bind_rows(
    taxation_gfs |> 
     filter(
       stringr::str_detect(revenue_name,"Stamp")),
    revenue_assessed |> 
      filter(
        stringr::str_detect(revenue_name,"Stamp")) |> 
      group_by(
        financial_year, state_name, revenue_name, revenue_type) |> 
      arrange(
        financial_year, state_name, revenue_name, 
        revenue_type, desc(update_year))|> 
      slice_head(n = 1)
    ) |> 
  left_join(population_erp, by = join_by("state_name","financial_year")) |>
  mutate(per_capita = (1e6*revenue)/population)


ggplot2::ggplot(
  data = stamp_per_capita,
  ggplot2::aes(
    x = financial_year,
    y = per_capita,
    colour = state_name,
    linetype = revenue_type)) + 
 ggplot2::geom_line(linewidth = .5) +
 ggplot2::scale_linetype_manual(
   name = NULL,
   values = c("Actual"=1,"Assessed"=2))+
 ggplot2::facet_wrap(
   facets = vars(state_name), nrow=2,
   labeller = 
     ggplot2::labeller(
       state_name = ggplot2::label_wrap_gen(width = 20)),
   axes = "all_y") +
  ggplot2::scale_colour_manual(
    guide = 'none',
    values = strayr::palette_state_name_2016,name=NULL)+
  ggplot2::coord_cartesian(
    xlim = c(lubridate::dmy("30-9-2018"),NA),
    ylim = c(0,2e3)
  ) +
  ggplot2::scale_x_date(
    name=NULL,
    date_labels = "%y",
    date_breaks = "1 years")+
  ggplot2::scale_y_continuous(name = "Revenue per capita, $ 000s") +
  ggplot2::guides(linetype = ggplot2::guide_legend(keywidth = 2)) +
  vpstheme::theme_vps_dh(base_size = 20)
```
